Reproducible, flexible and high throughput data extraction from primary literature: The \pkg{metaDigitise} \proglang{R} package

### 2017-11-20 docs/manuscript.tex

\section{Introduction}
In many different contexts, researchers need to make use of data presented in primary literature. Most notably, this includes meta-analysis, which is becoming increasingly common in many research fields. Meta-analysis uses effect size estimates and their sampling variance, taken from many studies, to understand whether particular effects are common across studies and to explain variation among these effects (Nakagawa et al. 2017). Meta-analysis therefore relies foremost on data extracted from primary literature and more specifically statistics that have been reported in the text or tables of research papers. These summary statistics are also, however, frequently presented in figures and so need to be manually extracted using digitising programs. Although there are several existing tools to perform tasks like this (e.g. DataThief (REF), GraphClick (REF), WebPlotDigitizer (REF))
, these tools are not designed this specific purpose (i.e. a meta-analysis). 
%% need to spell out clear points
%% 1 - dont manipulate data/allow different plot types
Firstly, they typically only provide the user with calibrated x,y coordinates from imported figures, and do not differentiate between common plot types that are use to present data. This means that a large amount of downstream data manipulation is subsequently required. For example, data are frequently presented in mean and error plots (Figure \ref{fig:plot_type}A), for which the user wants a mean and error estimate for each group presented in the figure. With existing programs, x,y coordinates of means and errors are returned, to which the user must manually discern between mean and error coordinates and assign points to groups. The error then needs to be calculated as the deviation from the mean, and then transformed to a standard deviation, depending on the type of error presented.
%% 2- adding metadata
Secondly, digitising programs do not easily allow the integration of metadata at the time of data extraction, such as  grouping points or adding variable names, group names and sample sizes.
%% 3 - not in statistical software
Additionally, existing stand-alone software does not allow easy import into commonly used statistical software (such as \proglang{R}), or try to optimize the research pipelines to facilitate editing and reproducibility. These are major issues because extracting from figures can be an incredibly time-consuming process. Furthermore, although meta-analysis is an important tool in consolidating the data from many studies, relying on transparency of those studies, many of the processes involved in data extraction are opaque and difficult to reproduce, making extending studies problematic. Having a tool that facilitates reproducibility in meta-analyses will increase transparency and go a long way to resolving the reproducibility crises we are seeing in many fields (REF).

>>>>

\section{Introduction}

In many different contexts, researchers need to make use of data presented in primary literature. Most notably, this includes meta-analysis, which is becoming increasingly common in many research fields. Meta-analysis uses effect size estimates and their sampling variance, taken from many studies, to understand whether particular effects are common across studies and to explain variation among these effects \citep{Naka2017, Glass1976, Koricheva2013, Borenstein2009}. Meta-analysis therefore relies foremost on data extracted from primary literature and more specifically statistics that have been reported in the text or tables of research papers. These summary statistics are also, however, frequently presented in figures and so need to be manually extracted using digitising programs. Although there are several existing tools to perform tasks like this (e.g. DataThief (REF), GraphClick (REF), WebPlotDigitizer (REF))
, these tools are not designed this specific purpose (i.e. a meta-analysis). 
%% need to spell out clear points
%% 1 - dont manipulate data/allow different plot types
Firstly, they typically only provide the user with calibrated x,y coordinates from imported figures, and do not differentiate between common plot types that are use to present data. This means that a large amount of downstream data manipulation is subsequently required. For example, data are frequently presented in mean and error plots (Figure \ref{fig:plot_type}A), for which the user wants a mean and error estimate for each group presented in the figure. With existing programs, x,y coordinates of means and errors are returned, to which the user must manually discern between mean and error coordinates and assign points to groups. The error then needs to be calculated as the deviation from the mean, and then transformed to a standard deviation, depending on the type of error presented.
%% 2- adding metadata
Secondly, digitising programs do not easily allow the integration of metadata at the time of data extraction, such as  grouping points or adding variable names, group names and sample sizes.
%% 3 - not in statistical software
Additionally, existing stand-alone software does not allow easy import into commonly used statistical software (such as \proglang{R}), or try to optimize the research pipelines to facilitate editing and reproducibility. These are major issues because extracting from figures can be an incredibly time-consuming process. Furthermore, although meta-analysis is an important tool in consolidating the data from many studies, relying on transparency of those studies, many of the processes involved in data extraction are opaque and difficult to reproduce, making extending studies problematic. Having a tool that facilitates reproducibility in meta-analyses will increase transparency and go a long way to resolving the reproducibility crises we are seeing in many fields \citep{peng_reproducible_2006, peng_reproducible_2011, sandve_ten_2013}.
